Mahotas: Open source software for scriptable computer vision

### 2012-07-05 paper.tex
TXT Start writing
I will switch to MIT License before releasing 1.0 and this paper

\section*{Abstract}
\section{Introduction}

Mahotas is a computer vision library for Python. It operates on numpy arrays,
but its inner loops are implemented in \cpp{} for speed and ease of
implementation.

@@

Often, functions take an argument named \code{output} where the output will be
stored. This argument is often much more restricted in type. In particular, it
must often be a contiguous array.\footnote{Numpy supports non-contiguous
arrays, which are most often slices into other, larger, contiguous arrays
(e.g., given a $128 \times 128$ contiguous array, one can build a $64 \times
128$ non-contiguous array by taking every other row).} Since this is a
performance feature, it is natural that the interface is less flexible
(accessing a contiguous array is much more efficient than a non-contiguous
one).

@@

The main reason that mahotas is in \cpp{} (and not in pure C) is to use templates.
Almost \cpp{} functionality is split across 2~functions:

\begin{enumerate}
\item A \code{py\_function} which uses the Python C~API to get arguments and
check them. This is almost always pure~C.
\item A template \code{function<dtype>} which works for the type \code{dtype}
performing the actual operation.
\end{enumerate}

@@

The template machinery is not that complicated and the functions using it are
very simple and easy to read. The only downside is that there is some expansion
of code size when the compiler instanciates the function for the several
integer and floating point types. Given the small size of these functions, this
is not a big issue.

@@

\begin{description}
\item[\code{gil\_release}] This is a ``resource-acquisition is object
initialisation'' (\textsc{raii})\footnote{\textsc{Raii} is a design pattern in
\cpp{}, or other languages with scope linked deterministic object destruction,
such as D, where a resource is represented by an object, whose constructor
acquires it and whose destructor releases it. This guarantees that the object
is correctly released even if the scope is left through an exception
\citep{Stroustrup1994}.} object that release the Python global interpreter lock
(\textsc{gil})\footnote{In the CPython interpreter, the most commonly used
implementation of Python, there is a global lock for many Python related
functionality, which limits parallelism.} in its constructor and gets it back
in its destructor. Normally, the template function will release the
\textsc{gil} after the Python-specific code is done.
\item[\code{array}] This is a thin wrapper around \code{PyArrayObject} that
knows its data type and has iterators which resemble the \cpp{} standard
library.
\item[\code{filter\_iterator}] This is taken from \code{scipy.ndimage} and it
is useful to iterate over an image and use a centered filter around each pixel
(it keeps track of all of the boundary conditions).
\end{description}

@@

\medskip

@@

Mahotas includes code ported and incorporated from other projects. In
particular, the \textsc{surf} implementation is a port from the code from
\textit{dlib},\footnote{Dlib's webpage is at \url{http://dlib.net}.} a very
good \cpp{} library by Davis King. I also gleaned some insight into the
implementation of these features from Christopher Evan's OpenSURF library and
its documentation \citep{evans2009}.\footnote{OpenSURF is available at
\url{http://www.chrisevansdev.com/computer-vision-opensurf.html}, where several
documents describe details of the implementation.} The code which interfaces
with the FreeImage library, was written by Zachary Pincus and some of the
support code was written by Peter J. Verveer for the \code{scipy.ndimage}
project. All of these contributions were integrated while respecting the
software licenses under which the original code had been released. Robert Webb,
a summer student, worked with me on the local binary patterns implementation.
Finally, I thank the several users who have reported bugs and submitted small
fixes and participated on the project mailing list.

>>>>

\section*{Abstract}
Mahotas is a computer vision library for Python. It contains traditional image
processing functionality such as filtering and morphological operations as well
as more modern computer vision functions for feature computation, including
local features.

The interface is in Python, a dynamic programming language, which is very
appropriate for fast development, but the algorithms are implemented in \cpp{}
and are tuned for speed.

Mahotas is available under a liberal open source license (MIT License) and is
available from \url{http://github.com/luispedro/mahotas} or the Python Package
Index (\url{http://pypi.python.org/pypi/mahotas}).

\section{Introduction}

Mahotas is a computer vision library for the Python. It operates on numpy arrays,
but its inner loops are implemented in \cpp{} for speed and ease of
implementation.

@@

Often, functions take an argument named \code{out} where the output will be
stored. This argument is often much more restricted in type. In particular, it
must often be a contiguous array.\footnote{Numpy supports non-contiguous
arrays, which are most often slices into other, larger, contiguous arrays
(e.g., given a $128 \times 128$ contiguous array, one can build a $64 \times
128$ non-contiguous array by taking every other row).} Since this is a
performance feature, it is natural that the interface is less flexible
(accessing a contiguous array is much more efficient than a non-contiguous
one).

@@

The main reason that mahotas is implemented in \cpp{} (and not in C, which is
the language of the Python interpreted) is to use templates. Almost \cpp{}
functionality is split across 2~functions:

\begin{enumerate}
\item A \code{py\_function} which uses the Python C~API to get arguments and
check them.
\item A template \code{function<dtype>} which works for the type \code{dtype}
performing the actual operation.
\end{enumerate}

@@

The template machinery makes the functions that use it very simple and easy to
read. The only downside is that there is some expansion of code size when the
compiler instanciates the function for the several integer and floating point
types. Given the small size of these functions, this is not a big issue.

@@

\begin{description}
\item[\code{gil\_release}] This is a ``resource-acquisition is object
initialisation'' (\textsc{raii})\footnote{\textsc{Raii} is a design pattern in
\cpp{}, or other languages with scope linked deterministic object destruction,
such as D, where a resource is represented by an object, whose constructor
acquires it and whose destructor releases it. This guarantees that the object
is correctly released even if the scope is left through an exception
\citep{Stroustrup1994}.} object that release the Python global interpreter lock
(\textsc{gil})\footnote{In the CPython interpreter, the most commonly used
implementation of Python, there is a global lock for many Python related
functionality, which limits parallelism.} in its constructor and gets it back
in its destructor. Normally, the template function will release the
\textsc{gil} after the Python-specific code is done. This allows several
mahotas functions to run concurrently.
\item[\code{array}] This is a thin wrapper around \code{PyArrayObject} that
knows its data type and has iterators which resemble the \cpp{} standard
library. This makes the code type-safer.
\item[\code{filter\_iterator}] This is taken from \code{scipy.ndimage} and it
is useful to iterate over an image and use a centered filter around each pixel
(it keeps track of all of the boundary conditions).
\end{description}

@@

\section{Discussion}

@@

Mahotas includes code ported and incorporated from other projects. In
particular, the \textsc{surf} implementation is a port from the code from
\textit{dlib},\footnote{Dlib's webpage is at \url{http://dlib.net}.} a very
good \cpp{} library by Davis King. I also gleaned some insight into the
implementation of these features from Christopher Evan's OpenSURF library and
its documentation \citep{evans2009}.\footnote{OpenSURF is available at
\url{http://www.chrisevansdev.com/computer-vision-opensurf.html}, where several
documents describe details of the implementation.} The code which interfaces
with the FreeImage library, was written by Zachary Pincus and some of the
support code was written by Peter J. Verveer for the \code{scipy.ndimage}
project. All of these contributions were integrated while respecting the
software licenses under which the original code had been released. Robert Webb,
a summer student at Carnegie Mellon University, worked with me on the initial
local binary patterns implementation. Finally, I thank the several users who
have reported bugs, submitted small fixes, and participated on the project
mailing list.

\textbf{Funding}: I was supported in my work by the Funda\c c\~{a}o para a
Ci\^encia e Tecnologia (grants SFRH/BD/37535/2007 and PTDC/SAU-GMG/115652/2008)
and by a grant from the Siebel Scholars Foundation.
